Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2024-04-17 analysis/paper/paper.tex
update pcj quarto extension

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11.5pt,
  leqno]{scrartcl}

@@

\usepackage[sfdefault]{roboto}
\usepackage{unicode-math}
\usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
\usepackage{indentfirst}

% Modify section headings
\usepackage{titlesec}
\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{}
\titleformat{\subsection}{\normalfont\bfseries}{}{.5em}{}
\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{\addperiod}
\newcommand{\addperiod}[1]{#1.} %https://tex.stackexchange.com/questions/137193/adding-a-period-after-section

\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt plus 1pt}
\linespread{1.1}

% Adjust caption style
\usepackage[labelfont=bf,
    labelsep=endash,
    singlelinecheck=off,
    format=plain,
    margin=1.5cm,
    aboveskip=12pt,
    belowskip=12pt,
    font=footnotesize,
    justification=justified]{caption}

% Defining document colors
\usepackage{xcolor}
\definecolor{darkgray}{HTML}{808080}
\definecolor{mediumgray}{HTML}{6D6E70}
\definecolor{ligthgray}{HTML}{d9d9d9}
\definecolor{pciblue}{HTML}{346E92}
\definecolor{opengreen}{HTML}{77933c}

%\usepackage[colorlinks=true,linkcolor=pciblue,urlcolor=pciblue,citecolor=pciblue]{hyperref} % Not working - links in pdf not changing colour. 
                                                         %Probably because hyperref is no the last package to be loaded, which it needs to be.
%\urlstyle{same}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={pciblue},
  filecolor={Maroon},
  citecolor={pciblue},
  urlcolor={pciblue},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{1.5}
\section{Introduction}\label{introduction}

Dental calculus has proven to be an excellent source of a wide variety
of information about our past. The increased accessibility and
advancement of methods in aDNA, paleoproteomics, and mass spectrometry,
has expanded our ability to identify biomarkers of diet and disease on
an increasingly large scale
(\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}; \citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014}).

One such collection of biomarkers is alkaloids, a plant-derived group of
compounds. Many alkaloids have important medicinal and psychoactive
effects in humans, and their direct detection, or detection of their
metabolites, is of great interest to archaeologists. Previous studies
have successfully recovered alkaloids in archaeological contexts,
including ceramics (\citeproc{ref-smithDetectionOpium2018}{Smith et al.,
2018}), pipes (\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et
al., 2012}), human hair
(\citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
2013}; \citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et
al., 2009}), and even dental calculus employing both targeted
(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}) and
untargeted approaches (\citeproc{ref-buckleyDentalCalculus2014}{Buckley
et al., 2014};
\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}). Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent
stability and ability to survive over long periods of time
(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
\citeproc{ref-tushinghamHuntergathererTobacco2013}{Tushingham et al.,
2013}).

@@

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002};
\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). This may also
allow various alkaloids of dietary and medicinal origin to become
incorporated in dental plaque. Dental plaque undergoes frequent
mineralisation events, ultimately causing the entrapped alkaloids and
their metabolites to become preserved within the dental calculus.
Barring intentional or accidental removal of the calculus during life,
burial, excavation, and post-excavation cleaning, the alkaloids can then
be detected by various methods to show a record of consumption during
life. Because drugs may be transferred from plasma to saliva, there is
often a close correlation between drugs detected in oral fluid and
blood, though there are differences in detected concentrations
(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007};
\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011};
\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009}). This was
also shown to be true for dental calculus and blood
(\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}),
making dental calculus a potentially useful substance for detecting
ancient alkaloids and other dietary compounds.

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour
(\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}),
and there are biological differences in dental calculus formation and
drug metabolism that are related to age and sex
(\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023};
\citeproc{ref-unoSexAgedependent2017}{Uno et al., 2017};
\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). The sample
consists of 27 males, 11 probable males, 2 probable females, and 1
female (Figure~\ref{fig-sample-demography}). We selected males due to a
higher occurrence of pipe notches and dental calculus deposits than
females (unpublished observation).

@@

\includegraphics{../figures/fig-sample-demography-1.pdf}

@@

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits
(\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994}). Age-at-death
was estimated using dental wear, auricular and pubic surface appearance,
cranial suture closure, and epiphyseal fusion
(\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002};
\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994};
\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985}), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

@@

\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}

@@

\includegraphics{../figures/fig-detection-preservation-1.pdf}

@@

\includegraphics{../figures/fig-polycorr-1.pdf}

@@

Theophylline and caffeine were positively correlated in our samples,
though to a lesser extent than nicotine and cotinine, so we are unable
to determine if they originated from the same source
(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
theophylline have very similar chemical structures, so we expect they
would experience similar rates of incorporation and degradation,
allowing us to interpret the ratio and correlations between the
compounds. Caffeine is present in coffee, tea, and cocoa beans, with
concentrations slightly higher in coffee
(\citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al., 2002};
\citeproc{ref-chinCaffeineContent2008}{Chin et al., 2008};
\citeproc{ref-srdjenovicSimultaneousHPLC2008}{Srdjenovic et al., 2008};
\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
Theophylline is present in both coffee beans and tea leaves, but in
negligible quantities
(\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
It is also a primary metabolite of caffeine produced by the liver. Given
the low correlation, there are likely multiple sources of caffeine and
theophylline in the population, with tea and coffee being the most
obvious.\\
Tea consumption had become widespread in the Netherlands by 1820,
reaching all parts of society
(\citeproc{ref-nierstraszTeaTrade2015}{Nierstrasz, 2015, p. 91}).
Historically, we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
special occasion occurring from 15.00-20.00 hours, where tea was served
along with the evening bread
(\citeproc{ref-schuijtemakerTeTheegasten2011}{Schuijtemaker, 2011}).
Many households also owned at least one coffee pot and tea pot
(\citeproc{ref-boumanBegravenis2017}{Bouman, 2017}). Distinguishing
between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is
present in higher quantities in chocolate compared to caffeine and
theophylline (\citeproc{ref-alanonAssessmentFlavanol2016}{Alañón et al.,
2016}; \citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al.,
2002}; \citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al.,
1988}). However, In addition to oral factors affecting alkaloid uptake
in dental calculus, there is some indication that theobromine does not
preserve well in the archaeological record
(\citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017}), and
frequent consumption of all three items would be difficult to parse.
Additionally, we do not understand well enough the effect of the burial
on these specific compounds, and the original concentration of the
compounds in plants can be quite variable
(\citeproc{ref-kingCautionaryTales2017}{King et al., 2017}).

Salicylic acid was found in all but one individual in our sample. It can
be extracted from the bark of willow trees, \emph{Salix alba}, and has
long been used for its pain-relieving properties
(\citeproc{ref-bruinsmaBijdragenTot1872}{Bruinsma, 1872, p. 119}). It is
also present in many plant-based foods
(\citeproc{ref-duthieNaturalSalicylates2011}{Duthie \& Wood, 2011};
\citeproc{ref-malakarNaturallyOccurring2017}{Malakar et al., 2017}),
including potatoes, which were a staple of the Beemsterpolder diet
(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012}). The extracted
quantity from our samples decreased over the three washes, followed by a
sharp increase in the final calculus extraction, which is what we would
expect to see if the salicylic acid was incorporated during life
(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
especially with salicylic acid, there is a possibility for the compound
to enter the calculus through contact with the surrounding soil.
Salicylic acid is a very mobile organic acid
(\citeproc{ref-badriRegulationFunction2009}{Badri \& Vivanco, 2009};
\citeproc{ref-chenCa2Dependent2001}{Chen et al., 2001}) and the
ubiqutous presence in our samples may be explained by the compound
leching into the dental calculus from the burial environment. We can
therefore not confidently rule out environmental contamination without
analysing samples from the surrounding soil.

@@

The only strictly modern compound (at least in a European context)
detected in the sample was cocaine, which was detected in the first
batch of samples. Our sample is derived from an early--mid 19th century
population, and cocaine was isolated in 1860 by Albert Niemann, and
entered popular medical practice in 1884. Coca arrived in Europe as
early as 1771, but as botanical specimens rather than for consumption,
and there were also issues importing enough viable specimens of coca for
cocaine extraction (\citeproc{ref-abucaCocaTrade2019}{Abduca, 2019, p.
108}; \citeproc{ref-mortimerHistoryCoca1901}{Mortimer, 1901, p. 179}).
This would have been the first case of coca-leaf-consumption in Europe;
however, we were unable to replicate any of the cocaine results in the
second batch. We suspect that the original detection of cocaine was a
result of lab contamination during analysis.

@@

Targeting individuals with moderate-to-large calculus deposits likely
biased our sample, as the presence of calculus may increase the risk of
premature death (\citeproc{ref-yaussyCalculusSurvivorship2019}{Yaussy \&
DeWitte, 2019}). Additionally, periodontal disease (often linked to the
presence of calculus) is a risk-factor for respiratory diseases, if
periodontal and respiratory pathogens enter the bloodstream
(\citeproc{ref-azarpazhoohSystematicReview2006}{Azarpazhooh \& Leake,
2006}; \citeproc{ref-scannapiecoRoleOral1999}{Scannapieco, 1999};
\citeproc{ref-scannapiecoPotentialAssociations2001}{Scannapieco \& Ho,
2001}). In our sample, the percentage of chronic maxillary sinusitis
(37.0\%) is lower than in another (more representative) male sample
(44.1\%) (\citeproc{ref-casnaUrbanizationRespiratory2021}{Casna et al.,
2021}), and the caries percentage is similarly lower in our sample
(17.6\%) than a more representative sample (22.9\%)
(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).\\
We used the presence/absence of a pipe notch and concurrent detection of
tobacco as a crude estimate of the accuracy of the method, which we
found to be around 59.3\%. This is a very rough estimate, as the
presence of a pipe notch is likely not a perfect indicator of whether or
not someone consumed tobacco. Dental calculus is also more transient
than for example bone, as it can become dislodged during life,
intentionally or unintentionally, eliminating all trace of the alkaloids
consumed prior to its removal.\\
Following burial, compound stability over time will play a large role,
as will microbial degradation of compounds by bacteria and fungi in soil
(\citeproc{ref-liuNicotinedegradingMicroorganisms2015}{Liu et al.,
2015}), as well as the soil environment, such as temperature, pH, and
oxygen availability (\citeproc{ref-lindholstLongTerm2010}{Lindholst,
2010}; \citeproc{ref-mackiePreservationMetaproteome2017}{Mackie et al.,
2017}).\\
Due to this, quantitation of the detected compounds may have limited
value in archaeological samples due to degradation, and will greatly
affect our correlations related to concentration. The detected quantity
of a compound will also depend on the quantity in dental calculus during
life, which is largely controlled by the quantity consumed, how often
the calculus was disrupted/removed, metabolic breakdown of the compound,
and inter- and intra-individual factors related to stages of biofilm
maturation (\citeproc{ref-lustmannScanningElectron1976}{Lustmann et al.,
1976}; \citeproc{ref-velskoMicrobialDifferences2019}{Velsko et al.,
2019}; \citeproc{ref-zijngeBiofilmArchitecture2010}{Zijnge et al.,
2010}). In short, this means it is not really possible to equate the
absence of a compound as evidence for the absence of consumption, which
complicates the interpretation of our results. We have attempted to
minimise errors occurring due to this limitation by including a
relatively large sample of individuals and replicating our analysis.
Although, given the relatively low detection rate seen in tobacco, this
remains a major limitation, and will likely be compounded by increasing
antiquity of the samples.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
  leqno]{scrartcl}

@@

\usepackage[sfdefault]{roboto}
\usepackage{unicode-math}
\usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
%\usepackage{indentfirst}

% Modify section headings
\usepackage[indentafter,explicit]{titlesec}
\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{#1}
%\titleformat{\section}{\normalfont\large\bfseries\centering}{\cdr@ifeq{\sectionname}{}{}{\sectionname~}\thesection.}{.5em}{#1}
\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection.}{.5em}{#1}
\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{#1. }

\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt plus 1pt}

% Adjust caption style
\usepackage[labelfont=bf,
    labelsep=endash,
    singlelinecheck=off,
    format=plain,
    margin=1.5cm,
    aboveskip=12pt,
    belowskip=12pt,
    %font=footnotesize,
    font={footnotesize,stretch=1},
    justification=justified]{caption}

% Defining document colors
\usepackage{xcolor}
\definecolor{PCIdark}{HTML}{346E92}
\definecolor{PCImedium}{HTML}{77A4BF}
\definecolor{PCIlight}{HTML}{9FBFD2}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={PCIdark},
  filecolor={Maroon},
  citecolor={PCIdark},
  urlcolor={PCIdark},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{1.1}
\section{Introduction}\label{introduction}

Dental calculus has proven to be an excellent source of a wide variety
of information about our past. The increased accessibility and
advancement of methods in aDNA, paleoproteomics, and mass spectrometry,
has expanded our ability to identify biomarkers of diet and disease on
an increasingly large scale
(\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014};
\citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}).

One such collection of biomarkers is alkaloids, a plant-derived group of
compounds. Many alkaloids have important medicinal and psychoactive
effects in humans, and their direct detection, or detection of their
metabolites, is of great interest to archaeologists. Previous studies
have successfully recovered alkaloids in archaeological contexts,
including ceramics (\citeproc{ref-smithDetectionOpium2018}{Smith et al.,
2018}), pipes (\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et
al., 2012}), human hair
(\citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et al.,
2009}; \citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
2013}), and even dental calculus employing both targeted
(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}) and
untargeted approaches (\citeproc{ref-buckleyDentalCalculus2014}{Buckley
et al., 2014};
\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}). Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent
stability and ability to survive over long periods of time
(\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
\citeproc{ref-tushinghamHuntergathererTobacco2013}{Tushingham et al.,
2013}; \citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}).

@@

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(\citeproc{ref-whiteDentalCalculus1997}{White, 1997};
\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002}). This may
also allow various alkaloids of dietary and medicinal origin to become
incorporated in dental plaque. Dental plaque undergoes frequent
mineralisation events, ultimately causing the entrapped alkaloids and
their metabolites to become preserved within the dental calculus.
Barring intentional or accidental removal of the calculus during life,
burial, excavation, and post-excavation cleaning, the alkaloids can then
be detected by various methods to show a record of consumption during
life. Because drugs may be transferred from plasma to saliva, there is
often a close correlation between drugs detected in oral fluid and
blood, though there are differences in detected concentrations
(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007};
\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009};
\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011}). This was also
shown to be true for dental calculus and blood
(\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}),
making dental calculus a potentially useful substance for detecting
ancient alkaloids and other dietary compounds.

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour
(\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}),
and there are biological differences in dental calculus formation and
drug metabolism that are related to age and sex
(\citeproc{ref-whiteDentalCalculus1997}{White, 1997};
\citeproc{ref-unoSexAgedependent2017}{Uno et al., 2017};
\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023}). The
sample consists of 27 males, 11 probable males, 2 probable females, and
1 female (Figure~\ref{fig-sample-demography}). We selected males due to
a higher occurrence of pipe notches and dental calculus deposits than
females (unpublished observation).

@@

\includegraphics{../figures/pcj-fig-sample-demography-1.pdf}

@@

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits
(\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994}). Age-at-death
was estimated using dental wear, auricular and pubic surface appearance,
cranial suture closure, and epiphyseal fusion
(\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985};
\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994};
\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002}),
and divided into the following categories: early young adult (18-24
years), late young adult (25-34 years), middle adult ( 35-49 years), old
adult (50+ years). Preservation was visually scored on a four-stage
scale (excellent, good, fair, poor) based on the surface condition of
the bones and the extent of taphonomic degradation.

@@

\includegraphics{../figures/pcj-fig-auth-plot-batch2-1.pdf}

@@

\includegraphics{../figures/pcj-fig-detection-preservation-1.pdf}

@@

\includegraphics{../figures/pcj-fig-polycorr-1.pdf}

@@

Theophylline and caffeine were positively correlated in our samples,
though to a lesser extent than nicotine and cotinine, so we are unable
to determine if they originated from the same source
(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
theophylline have very similar chemical structures, so we expect they
would experience similar rates of incorporation and degradation,
allowing us to interpret the ratio and correlations between the
compounds. Caffeine is present in coffee, tea, and cocoa beans, with
concentrations slightly higher in coffee
(\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988};
\citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al., 2002};
\citeproc{ref-srdjenovicSimultaneousHPLC2008}{Srdjenovic et al., 2008};
\citeproc{ref-chinCaffeineContent2008}{Chin et al., 2008}). Theophylline
is present in both coffee beans and tea leaves, but in negligible
quantities (\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et
al., 1988}). It is also a primary metabolite of caffeine produced by the
liver. Given the low correlation, there are likely multiple sources of
caffeine and theophylline in the population, with tea and coffee being
the most obvious.\\
Tea consumption had become widespread in the Netherlands by 1820,
reaching all parts of society
(\citeproc{ref-nierstraszTeaTrade2015}{Nierstrasz, 2015, p. 91}).
Historically, we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
special occasion occurring from 15.00-20.00 hours, where tea was served
along with the evening bread
(\citeproc{ref-schuijtemakerTeTheegasten2011}{Schuijtemaker, 2011}).
Many households also owned at least one coffee pot and tea pot
(\citeproc{ref-boumanBegravenis2017}{Bouman, 2017}). Distinguishing
between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is
present in higher quantities in chocolate compared to caffeine and
theophylline (\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et
al., 1988}; \citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et
al., 2002}; \citeproc{ref-alanonAssessmentFlavanol2016}{Alañón et al.,
2016}). However, In addition to oral factors affecting alkaloid uptake
in dental calculus, there is some indication that theobromine does not
preserve well in the archaeological record
(\citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017}), and
frequent consumption of all three items would be difficult to parse.
Additionally, we do not understand well enough the effect of the burial
on these specific compounds, and the original concentration of the
compounds in plants can be quite variable
(\citeproc{ref-kingCautionaryTales2017}{King et al., 2017}).

Salicylic acid was found in all but one individual in our sample. It can
be extracted from the bark of willow trees, \emph{Salix alba}, and has
long been used for its pain-relieving properties
(\citeproc{ref-bruinsmaBijdragenTot1872}{Bruinsma, 1872, p. 119}). It is
also present in many plant-based foods
(\citeproc{ref-duthieNaturalSalicylates2011}{Duthie \& Wood, 2011};
\citeproc{ref-malakarNaturallyOccurring2017}{Malakar et al., 2017}),
including potatoes, which were a staple of the Beemsterpolder diet
(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012}). The extracted
quantity from our samples decreased over the three washes, followed by a
sharp increase in the final calculus extraction, which is what we would
expect to see if the salicylic acid was incorporated during life
(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
especially with salicylic acid, there is a possibility for the compound
to enter the calculus through contact with the surrounding soil.
Salicylic acid is a very mobile organic acid
(\citeproc{ref-chenCa2Dependent2001}{Chen et al., 2001};
\citeproc{ref-badriRegulationFunction2009}{Badri \& Vivanco, 2009}) and
the ubiqutous presence in our samples may be explained by the compound
leching into the dental calculus from the burial environment. We can
therefore not confidently rule out environmental contamination without
analysing samples from the surrounding soil.

@@

The only strictly modern compound (at least in a European context)
detected in the sample was cocaine, which was detected in the first
batch of samples. Our sample is derived from an early--mid 19th century
population, and cocaine was isolated in 1860 by Albert Niemann, and
entered popular medical practice in 1884. Coca arrived in Europe as
early as 1771, but as botanical specimens rather than for consumption,
and there were also issues importing enough viable specimens of coca for
cocaine extraction (\citeproc{ref-mortimerHistoryCoca1901}{Mortimer,
1901, p. 179}; \citeproc{ref-abucaCocaTrade2019}{Abduca, 2019, p. 108}).
This would have been the first case of coca-leaf-consumption in Europe;
however, we were unable to replicate any of the cocaine results in the
second batch. We suspect that the original detection of cocaine was a
result of lab contamination during analysis.

@@

Targeting individuals with moderate-to-large calculus deposits likely
biased our sample, as the presence of calculus may increase the risk of
premature death (\citeproc{ref-yaussyCalculusSurvivorship2019}{Yaussy \&
DeWitte, 2019}). Additionally, periodontal disease (often linked to the
presence of calculus) is a risk-factor for respiratory diseases, if
periodontal and respiratory pathogens enter the bloodstream
(\citeproc{ref-scannapiecoRoleOral1999}{Scannapieco, 1999};
\citeproc{ref-scannapiecoPotentialAssociations2001}{Scannapieco \& Ho,
2001}; \citeproc{ref-azarpazhoohSystematicReview2006}{Azarpazhooh \&
Leake, 2006}). In our sample, the percentage of chronic maxillary
sinusitis (37.0\%) is lower than in another (more representative) male
sample (44.1\%) (\citeproc{ref-casnaUrbanizationRespiratory2021}{Casna
et al., 2021}), and the caries percentage is similarly lower in our
sample (17.6\%) than a more representative sample (22.9\%)
(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).\\
We used the presence/absence of a pipe notch and concurrent detection of
tobacco as a crude estimate of the accuracy of the method, which we
found to be around 59.3\%. This is a very rough estimate, as the
presence of a pipe notch is likely not a perfect indicator of whether or
not someone consumed tobacco. Dental calculus is also more transient
than for example bone, as it can become dislodged during life,
intentionally or unintentionally, eliminating all trace of the alkaloids
consumed prior to its removal.\\
Following burial, compound stability over time will play a large role,
as will microbial degradation of compounds by bacteria and fungi in soil
(\citeproc{ref-liuNicotinedegradingMicroorganisms2015}{Liu et al.,
2015}), as well as the soil environment, such as temperature, pH, and
oxygen availability (\citeproc{ref-lindholstLongTerm2010}{Lindholst,
2010}; \citeproc{ref-mackiePreservationMetaproteome2017}{Mackie et al.,
2017}).\\
Due to this, quantitation of the detected compounds may have limited
value in archaeological samples due to degradation, and will greatly
affect our correlations related to concentration. The detected quantity
of a compound will also depend on the quantity in dental calculus during
life, which is largely controlled by the quantity consumed, how often
the calculus was disrupted/removed, metabolic breakdown of the compound,
and inter- and intra-individual factors related to stages of biofilm
maturation (\citeproc{ref-lustmannScanningElectron1976}{Lustmann et al.,
1976}; \citeproc{ref-zijngeBiofilmArchitecture2010}{Zijnge et al.,
2010}; \citeproc{ref-velskoMicrobialDifferences2019}{Velsko et al.,
2019}). In short, this means it is not really possible to equate the
absence of a compound as evidence for the absence of consumption, which
complicates the interpretation of our results. We have attempted to
minimise errors occurring due to this limitation by including a
relatively large sample of individuals and replicating our analysis.
Although, given the relatively low detection rate seen in tobacco, this
remains a major limitation, and will likely be compounded by increasing
antiquity of the samples.
